---
title: "High performance GLM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{High performance GLM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hiperglm)
```

# Introduction

In this vignette, we give some examples on how to use the **hiperglm** package to perform linear regression on datasets.

We use the US Egg Production Data from TidyTuesday. The data contains two data frames containing the production of eggs and hens over time as well as the proportion of cage-free eggs and hens in the United States, which originates from **The Humane League** website
(url: <https://thehumaneleague.org/article/E008R01-us-egg-production-data>, <https://osf.io/z2gxn/>). The data dictionary can be found at the TidyTuesday Github repository <https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-04-11/readme.md>.

We aim to answer the question of how time and month of the year relates to the egg and hen production.

# Downloading the data

We store the data in a folder named **data**(and create the data beforehand if it does not exist already), and download the data only if no file of the same name exists in the folder. 
```{r, eval=FALSE}
if(!dir.exists('data')) system("mkdir data")
if(!'tidytuesdayR' %in% installed.packages()[,1])
  install.packages('tidytuesdayR')
library(tidytuesdayR)
tuesdata <- tidytuesdayR::tt_load(2023, week = 15)
eggproduction <- tuesdata$`egg-production`
cagefreepercentages <- tuesdata$`cage-free-percentages`
if(!file.exists('./data/eggprod.csv'))
write.csv(eggproduction,file='./data/eggprod.csv')
if(!file.exists('./data/cagefree.csv'))
write.csv(cagefreepercentages,file='./data/cagefree.csv')
```

Now we load the data from the downloaded file.
```{r}
eggprod <- read.csv('./data/eggprod.csv',row.names = 1)
cagefree <- read.csv('./data/cagefree.csv',row.names = 1)
```

# Data Wrangling and Visualization

We first take a glance at the original data.
```{r}
library(tidyverse)
head(eggprod)
attach(eggprod)
table(eggprod$prod_type)
table(eggprod$prod_process)
```
We will need to extract the years and months of each entry for further analysis, and create dummy variables for the 12 possible values(1-12) of months. Also, since the file source of each entry is irrelevent data, we need to remove it. Also, we need to rescale the number of hens and eggs for better numerical representation and computation. We rescale number of hens to be in millions, and number of eggs to be in billions.
```{r}
ep <- eggprod %>% 
  select(-source) %>%
  mutate(observed_month=as.Date(observed_month)) %>%
  mutate(year=year(observed_month), month=month(observed_month)) %>%
  mutate(n_hens=n_hens/1e6, n_eggs=n_eggs/1e9) %>%
  mutate(prod_type=as_factor(str_sub(prod_type,1,-6))) %>%
  rename(process=prod_process)
ep <- ep %>% 
  select(month) %>% 
  map_dfc(1:12,function(i,x) as.numeric(x==i),x=.) %>%
  set_names(nm=paste("Month",1:12,sep='')) %>%
  cbind(ep, .)
head(ep)
```
Now we perform some visualization on the data.
```{r}
p1 <- ep %>% ggplot() + 
  geom_line(aes(x=observed_month,y=n_hens,group=process,col=process)) +
  facet_grid(cols=vars(prod_type)) +
  xlab("Time") + ylab("Hen production(in millions)") +
  labs(title="Number of hens produced over time",
       subtitle="Grouped over production process and type",
       caption="Data is collected monthly")
p2 <- ep %>% ggplot() + 
  geom_line(aes(x=observed_month,y=n_eggs,group=process,col=process)) +
  facet_grid(cols=vars(prod_type)) +
  xlab("Time") + ylab("Egg production(in billions)") +
  labs(title="Number of eggs produced over time",
       subtitle="Grouped over production process and type",
       caption="Data is collected monthly")
p3 <- ep %>% ggplot() + geom_point(aes(x=n_hens,y=n_eggs,col=process)) + 
  geom_smooth(aes(x=n_hens,y=n_eggs,group=process,col=process),
              method='loess',formula='y~x',se=FALSE,linewidth=0.5) +
  scale_color_discrete() +
  xlab("Hen production(in millions)") + ylab("Egg production(in billions)") +
  labs(title="Relation of hen and egg production",
       subtitle="Grouped by production process",
       caption="Smoothing estimate included. Note that the different processes present similar patterns")
print(p1)
print(p2)
print(p3)
```
# Statistical Analysis

For the sake of demonstration, we assume that the hens production is the sum of a linear function of year, indicator of process variable, and a monthly offset for each month of the year. The eggs production is assumed to be a linear function of time, month indicators, interaction of time and process, and hen production, to include the apparent relationship between hens and eggs production.
```{r}
# Fit the model on hen production
fm <- paste(c("n_hens~year*process",paste("Month",2:12,sep='')),collapse="+",sep='')
X <- model.matrix(lm(as.formula(fm),ep))
y <- ep$n_hens
# Solve using Pseudo-inverse Algorithm(by Cpp)

hg_pinv <- hiper_glm(X,y,model='linear',option=list(solver='PINV'))

# Solve using R's OPTIM BFGS algorithm

hg_bfgs <- hiper_glm(X,y,model='linear',option=list(solver='BFGS'))

rbind(optim=coef(hg_bfgs),pinv=coef(hg_pinv))
```
```{r}
# Fit the model on egg production
fm <- paste(c("n_eggs~n_hens+year*process",paste("Month",2:12,sep='')),collapse="+",sep='')
X <- model.matrix(lm(as.formula(fm),ep))
y <- ep$n_eggs
# Solve using Pseudo-inverse Algorithm(by Cpp)

hg_pinv <- hiper_glm(X,y,model='linear',option=list(solver='PINV'))

# Solve using R's OPTIM BFGS algorithm

hg_bfgs <- hiper_glm(X,y,model='linear',option=list(solver='BFGS'))

rbind(optim=coef(hg_bfgs),pinv=coef(hg_pinv))
```
We can see that PINV and BFGS gives the same result. However, we can do further tests to see the speed difference between these two methods.
```{r}
system.time(replicate(1e5,hiper_glm(X,y,model='linear',option=list(solver='PINV'))))
system.time(replicate(1e5,hiper_glm(X,y,model='linear',option=list(solver='BFGS'))))
```
We can see that pseudo-inverse is slightly faster than BFGS in linear model solving.

# Summary

The hens production increases by 3.3 millions per year, while egg production is positively related to both time and hen production, and within each year, April and December shows peak production given that other conditions are fixed. Non-organic cage-free production increases faster over time while organic cage-free production increases slower than overall production.

# Functions Applied

**dplyr**: mutate, select

**lubridate**: month, year

**forcats**: as_factor

**stringr**: str_sub

**purrr**: map_dfc, set_names

**ggplot2**: geom_line, geom_point, geom_smooth, facet_grid

**hiperglm(package of interest)**: hiper_glm, coef.hglm(S3 generic of hglm class, the output of hiper_glm)